Two applications of logic programming to Coq

### 2020-11-03 papers/types2020/cultures.tex
comments to s 1 2 5

In many ways, proof theory is more elementary and low-level than most
approaches to type theory.  For example, type theories usually answer
the question ``What is a proof?'' with the response ``a (dependently)
typed $\lambda$-term''.  That is, when describing a type theory, one
usually decides that a proof is a certain kind of $\lambda$-term
within the system.  In contrast, the proof theory treats logical
propositions and proofs as separate.  For example, proof theory does
not assume that there are terms within the logic that describe proofs.

@@

Gentzen's discovery that the key to treating classical and
intuitionistic logics in the same proof system was the invention of
the weakening and contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This invention lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof that has not been given a satisfactory representation in
type theory.  This same innovation of Gentzen also opened the way to
another key proof-theoretic discovery, that of linear logic
\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
presentation of linear logic while most treatments of linear logic in
type theory have difficulty treating linear negation and the
multiplicative disjunction $\lpar$ and its unit $\bot$.

@@

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  Recent
developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
makes it possible to extract from sequent calculus not only natural
deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of Elpi with Coq.

@@

Given that \lP and Coq both result from a mixing of the
$\lambda$-calculus with logic, it is important to understand their
differences.  The confusion around the term \emph{higher-order
abstract syntax} (HOAS), for example, helps to illustrate their
differences.  In the functional programming setting, including the Coq
system, the HOAS approach leads to encoding binding structures within
terms using functions.  Such encodings are surprisingly complicated
since they allow for \emph{exotic terms} \cite{despeyroux95tlca} and
for structures on which induction is not possible
\cite{roeckl01fossacs}.  In the logic programming setting,
particularly \lP, HOAS is well supported since bindings are allowed
with terms ($\lambda$-bindings), in formulas (quantifiers), and in
proofs (eigenvariables).  For this reason, the term
\emph{$\lambda$-tree syntax} was introduced to name this particular
take on HOAS \cite{miller19jar}.  The Abella proof assistant
\cite{baelde14jfr} was designed, in part, to provide inductive and
co-inductive inference involving specifications using the
$\lambda$-tree syntax approach.

@@

While \lP is a typed language, it only uses simple, polymorphic types,
and these are more often used to indicate syntactic categories.  Any
dependency information needs to be captured using predicates and
quantifiers.

>>>>

In many ways, proof theory is more elementary and low-level than most
approaches to type theory.  For example, type theories usually answer
the question ``What is a proof?'' with the response ``a (dependently)
typed $\lambda$-term''.  That is, when describing a type theory, one
usually decides that a proof is a certain kind of $\lambda$-term
within the system.  In contrast,  proof theory treats logical
propositions and proofs as separate.  For example, proof theory does
not assume that there are terms within the logic that describe proofs.

@@

Gentzen's discovered that the key to treating classical and
intuitionistic logics in the same proof system was %the invention of
the weakening and contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This invention lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof that has not been given a satisfactory representation in
type theory.  This same innovation of Gentzen also opened the way to
another key proof-theoretic discovery, that of linear logic
\cite{girard87tcs}.  Similarly, sequent calculus provides an elegant
presentation of linear logic while most treatments of linear logic in
type theory have difficulty treating linear negation and the
multiplicative disjunction $\lpar$ and its unit $\bot$.

@@

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  
Developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
makes it possible to extract from sequent calculus not only natural
deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of Elpi with Coq.

@@

Given that \lP and Coq both result from combining the
$\lambda$-calculus with logic, it is important to understand some of
their their differences.  The confusion around the term
\emph{higher-order abstract syntax} (HOAS), is a case in point.  In
the functional programming setting, including the Coq system, the HOAS
approach leads to encoding binding structures within terms using
functions.  Such encodings are surprisingly complicated since they
allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
structures on which induction is not possible \cite{roeckl01fossacs}.
\begin{metanote}
  AM: I find this a bit unbalanced. Exotic terms can be avoided with
  Honsell's theory of context and/or Adam's PHOAS -- not that I like
  it.. And of course there is Beluga and friends.  On the other hand
  Coq users are happy with library-based approaches such as Autosubt or well-typed DB
  that allows for sphisticated treatments of binders. I would say something more generic such as that mixing functions and HOAS is not immediate, may require additional machinery both in terms of type theories than of code generation, while in lp\dots
\end{metanote}
In the logic programming setting, particularly \lP, HOAS is well
supported since bindings are allowed with terms ($\lambda$-bindings),
in formulas (quantifiers), and in proofs (eigenvariables).  For this
reason, the term \emph{$\lambda$-tree syntax} was introduced to name
this particular take on HOAS \cite{miller19jar}.  The Abella proof
assistant \cite{baelde14jfr} was designed, in part, to provide
inductive and co-inductive inference involving specifications using
the $\lambda$-tree syntax approach.

@@

While \lP is a typed language, it only uses simple, polymorphic types,
\begin{metanote}
AM:  simple are not poly: can we say ML-style polymorphism?
\end{metanote}
and these are more often used to indicate syntactic categories.  Any
dependency information needs to be captured using predicates and
quantifiers.

### 2020-11-03 papers/types2020/elab.tex
comments to s 1 2 5

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, the formula corresponding to
that type is, in fact, accepted by Coq users as a theorem of
intuitionistic logic.  The rest of the Coq system, especially its
tactic language, is designed to help a human user build
proofs-as-$\lambda$-terms that can be checked by the kernel.

There are many theorem provers for intuitionistic logic
\cite{raths07jar} for which a successful proof is not the kind of
detailed $\lambda$-term required by the Coq kernel.  Often, such
provers provide no information about the proofs they discover.  To the
extent that some prove evidence is output after a successful run, such
evidence is usually just a trace of some key aspects of a proof: some
details are often not included in such a trace.  For example:
\begin{enumerate}

@@

\item Check proofs by executing the logic programming checker that is
  parametrized by the particular FPC definition.
\end{enumerate}
%
It is also possible to pair two different proof certificates, defined
by two different FPC definitions, and do the proof checking in
parallel.  The logic programming setting allows such parallel checking
to also serve as checking-and-synthesizing.  That is, during the
checking of one certificate, it is possible to synthesize, for
example, a fully detailed term that is appropriate for handing to the
Coq kernel.

@@

To illustrate this approach to elaborating externally provided proof
structures into terms the Coq kernel can check, we consider here
examples take from the file \verb+coq_fpc.v+, which is part of the
repository \url{https://github.com/proofcert/fpc-elpi/}.
Figure~\ref{fig:debruijn} presents the specific FPC that defines the
proof semantics of using the De Briujn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.

@@

@@

>>>>

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, the formula corresponding to
that type is, in fact, accepted by Coq users as a theorem of
intuitionistic logic.
\begin{metanote}
  Small point: the Coq kernel can be soundly extended with axioms and will also validate classical tautologies
\end{metanote}
The rest of the Coq system, especially its
tactic language, is designed to help a human user build
proofs-as-$\lambda$-terms that can be checked by the kernel.

There are many theorem provers for intuitionistic logic
\cite{raths07jar} for which a successful proof is not the kind of
detailed $\lambda$-term required by the Coq kernel.  Often, such
provers provide no information about the proofs they discover.  To the
extent that some  evidence is output after a successful run, such
evidence is usually just a trace of some key aspects of a proof, where some
details are often not included.  For example:
\begin{enumerate}

@@

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
%
It is also possible to pair two different proof certificates, defined
by two different FPC definitions, and do the proof checking in
parallel.  The logic programming setting allows such parallel checking
to also serve as checking-and-synthesizing.  That is, during the
checking of one certificate, it is possible to synthesize, for
example, a fully detailed term that is appropriate for handing to the
Coq kernel.

@@

To illustrate this approach to elaborating externally provided proof
structures into terms the Coq kernel can check, we consider here
examples take from the file \verb+coq_fpc.v+, which is part of the
repository \url{https://github.com/proofcert/fpc-elpi/}.
Figure~\ref{fig:debruijn} presents the specific FPC that defines the
proof semantics of using the De Bruijn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.

@@

\begin{metanote}
  AM: I think we need more: the FPC may go to the appendix, we may want to say that there is a tactic that proves these theorems etc. 
\end{metanote}

@@

% LocalWords:  intuitionistic

### 2020-11-03 papers/types2020/intro.tex
comments to s 1 2 5

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the \texttt{coq-elpi} plugin to the Coq
proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
extends earlier implementations in primarily two directions:  First,
Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of Elpi comes equipped with  a quotation and anti-quotation syntax for mixing
Coq expressions with \lP program elements and an API for accessing the
Coq environment, typechecking included.

@@

Since relations (not functions) play a central role in \lP, Elpi is
capable to providing direct support for the many relations that have a
role in proof assistants implementation and usage.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics
specifications \cite{kahn87stacs,hannan93jfp}), and interaction (e.g.,
structured operational semantics \cite{plotkin81,milner89book}).

@@

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features,
% (e.g., a builtin treatment of bindings in syntax)
but also because it soundly
implements a higher-order intuitionistic logic.  The latter aspect of
Elpi makes it possible to encode the proofs and proof theory of
specific subsets of the Coq logic.

>>>>

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the \texttt{coq-elpi} plugin to the Coq
proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
extends earlier ones in primarily two directions:  First,
Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of Elpi comes equipped with  a quotation and anti-quotation syntax for mixing
Coq expressions with \lP program elements and an API for accessing the
Coq environment, typechecking included.

@@

Since relations (not functions) are central in \lP, Elpi is
capable to providing direct support for the many relations that have a
role in proof assistants implementation and usage.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics
specifications \cite{kahn87stacs,hannan93jfp}), and interaction (e.g.,
structured operational semantics \cite{plotkin81,milner89book}).

@@

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features,
% (e.g., a builtin treatment of bindings in syntax)
but also because it soundly
implements a higher-order intuitionistic logic.
\todo{Explain better? Our kernels are deep embedding, aren't they? -am}The latter aspect of
Elpi makes it possible to encode the proofs and proof theory of
specific subsets of the Coq logic.
